Advanced Data Structures - Succinct Bit Vector

### 2024-06-20 article/article.tex
add evaluation plot and data

@@

\title{Dummy title} %TODO Please add

@@

\author{Paul Hegenberg}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

@@

\acknowledgements{I want to thank \dots}%optional

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. 
\end{abstract}

\section{Typesetting instructions -- Summary}
\label{sec:typesetting-summary}

LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl. 
In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others. The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.

Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume. 
\paragraph*{Minimum requirements}

\begin{itemize}
\item Use pdflatex and an up-to-date \LaTeX{} system.
\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
\item Use a spellchecker to correct typos.
\end{itemize}

\paragraph*{Mandatory metadata macros}
Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
\begin{itemize}
\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
\item \verb+\keywords+ (a comma-separated list of keywords).
\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
\item \verb+\begin{abstract}...\end{abstract}+ .
\end{itemize}

\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
Generally speaking, please do not override the \texttt{lipics-v2021}-style defaults. To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2021} defaults}. In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
\begin{itemize}
\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
\item Do not alter the spacing of the \texttt{lipics-v2021.cls} style file.
\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
 \verb+\begin{enumerate}[(a)]+ or the like.)
\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions. 
\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
\item Do not change the bibliography style. In particular, do not use author-year citations. (The
\texttt{natbib} package is not supported.)
\end{itemize}

\enlargethispage{\baselineskip}

This is only a summary containing the most relevant details. Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2021} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
\href{http://drops.dagstuhl.de/styles/lipics-v2021/lipics-v2021-authors/lipics-v2021-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2021/\newline lipics-v2021-authors/lipics-v2021-authors-guidelines.pdf}}

\section{Lorem ipsum dolor sit amet}

Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. Donec eget odio et magna ullamcorper vehicula ut vitae libero. Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis. Pellentesque id ante erat. In hac habitasse platea dictumst. Curabitur a scelerisque odio. Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam. Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.

\begin{lemma}[Lorem ipsum]
\label{lemma:lorem}
Vestibulum sodales dolor et dui cursus iaculis. Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper. Proin facilisis gravida rutrum. Etiam sed sollicitudin lorem. Proin pellentesque risus at elit hendrerit pharetra. Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
\end{lemma}

\begin{proof}
Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.

\proofsubparagraph*{Just some paragraph within the proof.}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\begin{claim}
content...
\end{claim}
\begin{claimproof}
content...
    \begin{enumerate}
        \item abc abc abc \claimqedhere{}
    \end{enumerate}
\end{claimproof}

\end{proof}

\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
\label{lemma:curabitur}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\end{corollary}

\begin{proposition}\label{prop1}
This is a proposition
\end{proposition}

\autoref{prop1} and \cref{prop1} \ldots

\subsection{Curabitur dictum felis id sapien}

Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat. Curabitur sed velit diam. Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu. Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet. Duis rutrum dapibus ornare. Nulla vehicula vulputate iaculis. Proin a consequat neque. Donec ut rutrum urna. Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Aenean nec faucibus leo. Cras ut nisl odio, non tincidunt lorem. Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat. Fusce risus libero, convallis at fermentum in, dignissim sed sem. Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}. Donec non suscipit lorem. Nam sit amet enim vitae nisl accumsan pretium.

\begin{lstlisting}[caption={Useless code.},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
for i:=maxint to 0 do 
begin 
    j:=square(root(i));
end;
\end{lstlisting}

\subsection{Proin ac fermentum augue}

Proin ac fermentum augue. Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis. Nulla facilisi. Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac. Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna. Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui. Praesent a pretium quam. Mauris non mauris sit amet eros rutrum aliquam id ut sapien. Nulla aliquet fringilla sagittis. Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor. Nulla cursus aliquet enim. Cras sapien risus, accumsan eu cursus ut, commodo vel velit. Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel. Integer faucibus faucibus felis.

\begin{itemize}
\item Ut vitae diam augue. 
\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem. 
\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit. 
\end{itemize}

\begin{remark}
content...
\end{remark}

\section{Pellentesque quis tortor}

Nec urna malesuada sollicitudin. Nulla facilisi. Vivamus aliquam tempus ligula eget ornare. Praesent eget magna ut turpis mattis cursus. Aliquam vel condimentum orci. Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi. Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.

\begin{lemma} [Quisque blandit tempus nunc]
Sed interdum nisl pretium non. Mauris sodales consequat risus vel consectetur. Aliquam erat volutpat. Nunc sed sapien ligula. Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus. Nunc vestibulum nunc ac massa pretium pharetra. Nulla facilisis turpis id augue venenatis blandit. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
\end{lemma}

Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.

\section{Morbi eros magna}

Morbi eros magna, vestibulum non posuere non, porta eu quam. Maecenas vitae orci risus, eget imperdiet mauris. Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis. Sed condimentum convallis dolor, a dignissim est ultrices eu. Donec consectetur volutpat eros, et ornare dui ultricies id. Vivamus eu augue eget dolor euismod ultrices et sit amet nisi. Vivamus malesuada leo ac leo ullamcorper tempor. Donec justo mi, tempor vitae aliquet non, faucibus eu lacus. Donec dictum gravida neque, non porta turpis imperdiet eget. Curabitur quis euismod ligula.

@@

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

List of different predefined enumeration styles:

\begin{itemize}
\item \verb|\begin{itemize}...\end{itemize}|
\item \dots
\item \dots
%\item \dots
\end{itemize}

\begin{enumerate}
\item \verb|\begin{enumerate}...\end{enumerate}|
\item \dots
\item \dots
%\item \dots
\end{enumerate}

\begin{alphaenumerate}
\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
\item \dots
\item \dots
%\item \dots
\end{alphaenumerate}

\begin{romanenumerate}
\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
\item \dots
\item \dots
%\item \dots
\end{romanenumerate}

\begin{bracketenumerate}
\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
\item \dots
\item \dots
%\item \dots
\end{bracketenumerate}

\begin{description}
\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
\item[Description 2] Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\item[Description 3]  \dots
%\item \dots
\end{description}

\cref{testenv-proposition} and \autoref{testenv-proposition} ...

\section{Theorem-like environments}\label{sec:theorem-environments}

List of different predefined enumeration styles:

\begin{theorem}\label{testenv-theorem}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{theorem}

\begin{lemma}\label{testenv-lemma}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{lemma}

\begin{corollary}\label{testenv-corollary}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{corollary}

\begin{proposition}\label{testenv-proposition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proposition}

\begin{conjecture}\label{testenv-conjecture}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{conjecture}

\begin{observation}\label{testenv-observation}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{observation}

\begin{exercise}\label{testenv-exercise}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{exercise}

\begin{definition}\label{testenv-definition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{definition}

\begin{example}\label{testenv-example}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{example}

\begin{note}\label{testenv-note}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note}

\begin{note*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note*}

\begin{remark}\label{testenv-remark}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark}

\begin{remark*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark*}

\begin{claim}\label{testenv-claim}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim}

\begin{claim*}\label{testenv-claim2}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim*}

\begin{proof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proof}

\begin{claimproof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claimproof}

>>>>

\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{statistics, units, groupplots}

\pgfplotsset{
    only if/.style args={entry of #1 is #2}{
        /pgfplots/boxplot/data filter/.code={
            \edef\tempa{\thisrow{#1}}
            \edef\tempb{#2}
            \ifx\tempa\tempb
            \else
                \def\pgfmathresult{}
            \fi
        }
    }
}

\usepackage{tikz}
\usetikzlibrary[fadings]

\def\code#1{\texttt{#1}}
\newcommand\prevsum[0]{\ensuremath{\overset{\leftarrow}{\Sigma}_1}}

@@

\title{Advanced Data Structures - Succinct Bit Vector}

@@

\author{Paul Hegenberg}{Karlsruher Institut für Technologie, Germany \and \url{https://blog.hegenberg.dev} }{urlgh@student.kit.edu}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\authorrunning{P. Open Access and P. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Paul Open Access} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{Bit Vectors} %TODO mandatory; please add comma-separated list of keywords

% \category{} %optional, e.g. invited paper

% \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

@@

% \acknowledgements{I want to thank \dots}%optional

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
\end{abstract}

\section{Algorithm}
In the following section we will present two main data structures which implement the bit vector
operations

\subsection{Interleaved}
The interleaved data structure aims to use the CPU cache as efficiently as possible, where both
vector information and metadata should fit in the same cache line of 512 bit.
Our bit vector implementation consists of \code{uint64} blocks, of which we could fit 8 inside one cache line.
To speed up the computation of rank and select we need metadata, which we store in the first
64 bits of the line. This data structure is depicted in Figure~\ref{fig:interleaved}. 
This leaves us with a memory overhead of $12.5\%$. The used metadata $\prevsum$ is an incremental sum of previous bits
set to one. For an efficient build time of the data structure we make use of the pop count $c(\vec{v})$
instruction, which counts the number of 1 bits in a subvector $\vec{v}$, in our case an \code{uint64} number.

\begin{figure}[htbp]
    \centering
    \tikzfading[name=fade right,
            left color=transparent!0,
            right color=transparent!100]
        \tikzfading[name=fade left,
            left color=transparent!100,
            right color=transparent!0]
    \begin{tikzpicture}

        \draw[draw opacity=.4,text opacity=.4,path fading=fade left,densely dotted] (-3,0) rectangle node {$\vec{v}_6$} ++(2,1)
            ++ (0,-.5) -- ++(1,0);

        \draw (0,1) rectangle node {$\prevsum$} (2,0) 
            ++ (0,0) rectangle node {$\vec{v}_0$} ++(2,1)
            ++ (0,-.5) -- node[label={\dots}] {} ++(2,0)
            ++ (0,-.5) rectangle node {$\vec{v}_6$} ++(2,1);

        \draw[|-|] (0,-.3) -- node[label=below:{Vector Line (512 bit)}] {} ++(8,0);
        \draw[|-|] (0,1.3) -- node[label=above:{64 bit}] {} ++(2,0);

        \draw[draw opacity=.4,text opacity=.4,path fading=fade right,densely dotted] (8,.5) -- ++(1,0) 
            ++(0,-.5) rectangle node {$\prevsum$} ++(2,1);

    \end{tikzpicture}
    \caption{Interleaved Data Structure}
    \label{fig:interleaved}    
\end{figure}

\subsubsection{Rank}
The \code{rank} operation is defined as $\text{rank}(\alpha, p) \mapsto n$, where $\alpha \in \{0,1\}$
and $p,n \in \mathbb{N}_0$. Where $n$ is the number of $\alpha$ valued bits set before position $p$.
To calculate $n$ for a given $\alpha, p$ we first have to calculate the vector line where the Prev. Sum
is stored. Then we can use bit masks and the pop count operation to count the missing 1 bits.
$$
    n' = \prevsum + \sum_{i=0}^{j-1}c(\vec{v}_i) + c(\text{masked}(\vec{v}_j)) 
$$
We need to execute 7 pop count operations at most, but since everything fits into one cache line
we don't need to load additional information from memory. With $\alpha = 1: n = n'$ for $\alpha = 0: n = p - n'$.

\subsubsection{Select}
The \code{select} operation is defined as $\text{select}(\alpha, n) \mapsto p$, where $p$ is the position
of the $n$th $\alpha$ valued bit. With the interleaved data structure we use the incremental sum
to find the vector line with the biggest prev. sum $< n$ using a binary search.
The following vectors $\vec{v}_i$ will therefore contain the $n$th $\alpha$ bit.
To find the position we conduct a linear search utilizing pop count, selecting the first
$\vec{v}_i$ where $c(\vec{v}_i) \geq n$. Next we divide $\vec{v}_i$ into byte sized blocks to apply 
the same linear search to find the byte containing the $n$th bit.
To get the position inside the byte we created a lookup table, which is described further in Section~\ref{p:select_lookup}.

\paragraph{Select lookup table}
\label{p:select_lookup}
For a fast select lookup we created a lookup table, consisting of 8 columns for each bit,
as well as 256 rows for each possible byte.
The $x$-th column in the $y$-th row stores the position of the $x$-th bit in $y$.
Since most numbers don't contain 8 one bits, we added a validation bit $v$.
If $v = 1$ the stored position is valid. For each possible byte we need to store 32 bits.
A schematic view of said lookup table is depicted in Figure~\ref{fig:select_lookup}.
As this information is constant we included the table as a constant array in the compiled binary.

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}
        \draw[|-|] (0,1.5) -- node[label=above:32 bit] {} ++(9.5,0);

        \foreach \x in {1,...,3} \draw (\x*2.25-1.25,1) node {\x};

        \draw (7,1) node {$\dots$} ++ (1.5,0) node {8};

        \foreach \x in {0,...,1,3}
        \draw (0,-\x*.75) rectangle node {v} ++(.5,.5) ++(0,-.5) rectangle node {3 bit} ++(1.5,.5)
            ++(.25,-.5) rectangle node {v} ++(.5,.5) ++(0,-.5) rectangle node {3 bit} ++(1.5,.5)
            ++(.25,-.5) rectangle node {v} ++(.5,.5) ++(0,-.5) rectangle node {3 bit} ++(1.5,.5)
            ++(0,-.25) -- node {} ++(1,0)
            ++(0,-.25) rectangle node {v} ++(.5,.5) ++(0,-.5) rectangle node {3 bit} ++(1.5,.5);

        % \draw (0,3) rectangle node {v} ++(.5,.5) ++(0,-.5) rectangle node {3 bit} ++(1.5,.5);

        \draw (-.5,.25) node {0} 
            ++(0,-.75) node {1}
            ++(0,-.75) node {$\vdots$}
            ++(0,-.75) node {255};

    \end{tikzpicture}
    \caption{Byte select lookup table}
    \label{fig:select_lookup}
\end{figure}

\section{Evolution}

\section{Evaluation}
In this chapter we will evaluate our predicted runtime against measurements.
For the pre-computation we calculate an incremental sum $\prevsum$ every 7 blocks of 64 bits.
This should scale linearly with the vector size $|\vec{V}|=n$.
As for \code{rank} we can calculate the position of the vector line in constant time.
For \code{select} we need to first run a binary search in $\mathcal{O}(\log{n})$ to find the vector line.
After that we compute a linear search in $\mathcal{O}(7)$ as there is a maximum of 7 blocks we need to search.
The runtime will be dominated by the binary search after a certain time.
In Figure~\ref{fig:performance_eval} we can validate these predictions.

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}    
        \begin{axis}[
            legend columns=3,
            width=\textwidth*0.9,
            height=7cm,
            ymode=log,
            log basis y=10,
            ylabel={runtime},
            xlabel={vector size},
            y SI prefix=milli,
            y unit=s,
            every axis legend/.append style={
                at={(0.5,1.03)},
                anchor=south
            }, 
            xmin=28,
            xticklabel={$2^{\pgfmathparse{\tick}\pgfmathprintnumber{\pgfmathresult}}$},
            enlarge x limits=auto,
            grid=both,
            tick align=outside,
            tickpos=left,
        ]
            \addplot [
                mark=+,
                error bars/.cd,
                    y dir=both,
                    y explicit,
            ] table [x=bits,y=mean,y error=error] {errorplot.dat};
            \addlegendentry{Runtime}

            \addplot [
                mark=+,
                color=red,
                error bars/.cd,
                    y dir=both,
                    y explicit,
            ] table [x=bits,y=commandMean,y error=commandError] {errorplot.dat};
            \addlegendentry{Command Execution}

            \addplot [
                mark=+,
                color=blue,
                error bars/.cd,
                    y dir=both,
                    y explicit,
            ] table [x=bits,y=precomMean,y error=precomError] {errorplot.dat};
            \addlegendentry{Precomputation}
        \end{axis} 
    \end{tikzpicture}
    \caption{Performance Evaluation of an increasing vector size and a constant 100 000 commands.}
    \label{fig:performance_eval}
\end{figure}

% \begin{tikzpicture}
% 	\begin{axis}[
%         boxplot,
% 		boxplot/draw direction=y,
% 		ylabel={runtime},
%         y SI prefix=milli,y unit=s,
%         xmin=0,
%         xticklabel={$2^{\pgfmathparse{8+\tick}\pgfmathprintnumber{\pgfmathresult}}$},
%         % x tick label as interval,
%         xlabel={vector length},
% 	]

%     % \addplot table[y=time, only if={entry of bits is 31}]{benchmark.dat}; 
%     % \addplot table[y=time, only if={entry of bits is 32}]{benchmark.dat}; 
%     \pgfplotsinvokeforeach{8,...,32}{ 
%         \addplot table[y=time,x=bits,only if={entry of bits is #1}]{benchmark.dat}; 
%         % \addlegendentry{Cond. $2^{#1}$};
%     }

% 	\end{axis}
%     \begin{axis}[
%         boxplot,
% 		boxplot/draw direction=y,
% 		ylabel={runtime},
%         y SI prefix=milli,y unit=s,
%         xmin=0,
%         xticklabel={$2^{\pgfmathparse{8+\tick}\pgfmathprintnumber{\pgfmathresult}}$},
%         % x tick label as interval,
%         xlabel={vector length},
% 	]

%     % \addplot table[y=time, only if={entry of bits is 31}]{benchmark.dat}; 
%     % \addplot table[y=time, only if={entry of bits is 32}]{benchmark.dat}; 
%     \pgfplotsinvokeforeach{8,...,32}{ 
%         \addplot table[y=time,x=bits,only if={entry of bits is #1}]{benchmark.dat}; 
%         % \addlegendentry{Cond. $2^{#1}$};
%      }

%     \end{axis}
% \end{tikzpicture}

@@
